---
output:
  md_document:
    variant: markdown_github
---

[![Travis-CI Build Status](https://travis-ci.org/mdsumner/spbabel.svg?branch=master)](https://travis-ci.org/mdsumner/spbabel)

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE,message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figure/README-"
)
library(dplyr)
```

# spbabel: bisque or bouillon?

Part of an effort towards dplyr/table-fication of Spatial classes and the spatial-zoo in R. Inspired by Eonfusion and Manifold and dplyr and Douglas Adams and spurned on by helpful twitter and ropensci discussions. 


## Installation

Spbabel can be installed directly from github: 

```{r,eval=FALSE}
devtools::install_github("mdsumner/spbabel")
```

# What is the spatial babel for? 

Spatial data in the `sp` package have a formal definition  (extending class `Spatial`) that is modelled on shapefiles, and close at least in spirit to the Simple Features definition. https://github.com/edzer/sfr

In GIS we have Objects linked to a table of attributes, one row of a table for each Object. An Object is a shape defined by potentially complex *geometry*, and importantly the geometry is an *inherent* property of the shape while the attribute data is simply 
Internally Spatial objects are composed of lists of lists of matrices, with two levels of structure for the geometry for each each line or polygon object. An individual matrix is an ordered set of connected vertices defining a single linear path. If the path is closed it is a polygon "ring", and the order of the coordinates (clockwise or anti-clockwise reflects its status as an "island" or a "hole"). The formal base-level classes `Line` and `Polygon` store this path for the corresponding type, (these types cannot be mixed in a composite object).  An object may be composed of one or more linear paths of either type, and these are collected together as `Lines` and `Polygons`. So, a single `Line` or `Polygon` corresponds to a piece (connected line, island or hole), and a single `Lines` or `Polygons` object corresponds to  The formal definition of 




## Embedded nesting

Still working on the terminology here, does "embedding" make sense? 

This approach stores a single table at the top level, but it only stores the tables that we need - it's not a recursive object with a table split across rows. This means we can still normalize across objects (unique vertices), and apply operations on the entire set of rows (e.g. all vertices) without unnesting or recursing.  Yes this also means shared topology. 

Here is an example of "embedding". 

```{r}
library(spbabel)
library(maptools)
data(wrld_simpl)
world <- db_df(wrld_simpl)

world$Table

```

How to reconstitute these objects? 

```{r}
nam <- c("Australia", "New Zealand")
library(dplyr)
Objects <- function(x) x$Table$Object
Branches <- function(x) x$Table$Branch
Coords <- function(x) x$Table$Coord

(countryObjects <- Objects(world) %>% select(NAME, object) %>% filter(NAME %in% nam) )

countries <- countryObjects %>% inner_join(Branches(world)) %>% inner_join(Coords(world))

spFromTable(countries, attr = countryObjects, quiet = TRUE)
```

## Why do this? 

I want these things: 

* flexibility in the number and type/s of attribute stored as "coordinates", x, y, lon, lat, z, time, temperature, etc.
* shared vertices
* ability to store points, lines and areas together, sharing topology where appropriate


## Recursive nesting


I'm not particularly fond of this approach, since you need to nest two levels down in order to have tables of coordinates for parts, and tables of parts for objects. This means we start reproducing the recursive list structures we started with. Better is to store separate tables for distinct entities and keep rules for table views.

Recent discussion resulted in a "nested data frame" approach, I've applied that in `nested.r` with two-levels of nesting, Objects as the table to store each Branch (part or piece or ring or linestring), and each Branch table stores actual vertices (x, y). 

This example shows raw round-tripping, just to prove it mostly works. 

```{r}
library(tidyr)
library(spbabel)
library(maptools)
data(wrld_simpl)
xa <- nest(wrld_simpl)
## method for plotting this nsp_df (nested sp df) shows round-trip back to Spatial
plot(xa, col = grey(seq(0, 1, length = nrow(xa))))

## ggplot2
library(ggplot2)
library(tidyr)

## custom stat for this nested "nsp_df" object
ggplot(xa) + stat_nested()
 
```

This example shows we can pipeline in simple ways. 

Filter out by country NAME. 

```{r}
library(dplyr)

xa %>% filter(NAME %in% c("Australia", "New Zealand")) %>% 
 ggplot() + stat_nested()

```

Extract the full vertices table (with all object attributes dropped), filter-join on any objects that have holes and pick one to plot. 

(Here the cascade of un-nesting should extract the branch-level metadata, i.e. winding order, hole-status, area, etc.)

Using `geom_holygon` we can get true polypaths. 

```{r}
xa %>% inner_join(spbabel:::vertices(xa) %>% filter(hole == 1) %>% distinct(object)  %>% select(object)) %>%  
  filter(row_number() == 17) %>% ggplot() + geom_holygon() 


sptable(subset(wrld_simpl, NAME == "Armenia"))
armenia <- xa %>% filter(NAME == "Armenia") 

ggplot(armenia, aes(x, y, group = branch, fill = factor(hole))) + stat_nested()
```


Much more to do . . . 


```{r}
## how to pass down object for faceting . . .
#xa %>% inner_join(spbabel:::vertices(xa) %>% filter(hole == 1) %>% distinct(object)  %>% select(object)) %>%  
#   ggplot() + stat_nested() + facet_grid(branch)



```



Have a closer look at what happens, we need to keep at least `Object`, and we copy out `ISO3` to all vertices. This is essentially an implicit join, made simpler since rows store data frames 'recursively'. 

```{r}
library(dplyr)
library(tidyr)

## Be careful to keep the nested Object table - here see all vertices
xa %>% select(Object, ISO3) %>% unnest() %>% unnest()

```

Questions: how do we protect special columns, like "Object" - and how do we ensure they are recorded when unnested? Same goes for PROJ.4 string, here I use an attribute. 

I also tried store list of Polygons() in the column, that's kind of simple but you need S3 methods for everything - this is is `sp-df.r`. 

Nested data frames are much easier, I find it more natural to use two-level nesting, but you could go further to store more information on individual pieces, their area, winding, etc. You could also just store the fortify-table at one level nesting, more to experiment with. 

*See gris for deduplication of vertices - we cannot do that with nesting afaics.*


## Usage

Apply pipeline modifications to the attribute data of `sp` objects with dplyr verbs. 

```{r}
data(quakes)
library(sp)
coordinates(quakes) <- ~long+lat
library(spbabel)
## plot a subset of locations by number of stations
quakes %>% dplyr::filter(mag <5.5 & mag > 4.5) %>% select(stations) %>% spplot

```

We can use polygons and lines objects as well. 

```{r}
library(maptools)
data(wrld_simpl)

x <- wrld_simpl %>% mutate(lon = coordinates(wrld_simpl)[,1], lat = coordinates(wrld_simpl)[,2]) %>% filter(lat < -20, lon > 60) %>% select(NAME)
plot(x, asp = ""); text(coordinates(x), label = x$NAME, cex = 0.6)
```

use the `sptable<-` replacment method to modify the underlying geometric attributes (here `x` and `y` is assumed no matter what coordinate system). 

```{r}
## standard dplyr on this S4 class
w2 <- filter(wrld_simpl, NAME == "Australia")
plot(w2, col = "grey")

## modify the geometry on this object
sptable(w2) <- sptable(w2) %>% mutate(x = x - 5)

plot(w2, add = TRUE)
```

We can also restructure objects. 

```{r}
## explode (disaggregate) objects to individual polygons
## here each part becomes and object, and each object only has one part
w3 <- spFromTable(sptable(w2)  %>% mutate(object = part, part = 1), crs = proj4string(w2))
```



## TODO

Create idioms for modifying the geometry with dplyr verbs and/or piping. 

Consider whether `summarise` is a sensible for Spatial. 

Can we do geometry manipulations like `sptable(w2) <- sptable(w2) %>% mutate(object = part, part = 1)` and sensibly restore the attributes on the way? Just copy the value outs?


Implement `sample_n` and `sample_frac`. 

Implement the joins. 


## Approach

Create methods for the dplyr verbs: filter, mutate, arrange, select etc. 

This is part of an overall effort to normalize Spatial data in R, to create a system that can be stored in a database. 

Functions `sptable` and `spFromTable` create `tbl_df`s from Spatial classes and round trip them. This is modelled on `raster::geom rather` than `ggplot2::fortify`, but perhaps only since I use raster constantly and ggplot2 barely ever. 

Complicating factors are the rownames of sp and the requirement for them both on object IDs and data.frame rownames, and the sense in normalizing the geometry to the table of coordinates without copying attributes. 

(See `mdsumner/gris` for normalizing further to unique vertices and storing parts and objects and vertices on different tables. Ultimately this should all be integrated into one consistent approach.)



Thanks to @holstius for prompting a relook under the hood of dplyr for where this should go: https://gist.github.com/holstius/58818dc9bbb88968ec0b

This package `spbabel` and these packages should be subsumed partly in the overall scheme: 

https://github.com/mdsumner/dplyrodbc
https://github.com/mdsumner/manifoldr
