---
output: github_document
---

[![Travis-CI Build Status](https://travis-ci.org/mdsumner/spdplyr.svg?branch=master)](https://travis-ci.org/mdsumner/spdplyr)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/mdsumner/spdplyr?branch=master&svg=true)](https://ci.appveyor.com/project/mdsumner/spdplyr)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/spdplyr)](https://cran.r-project.org/package=spdplyr)
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE,message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "figure/README-"
)
library(maptools)
library(spdplyr)
```

# spdplyr

`spdplyr` provides methods for the `dplyr` verbs for the `Spatial` classes in the `sp` package. 


## Installation

Spdplyr can be installed directly from Github: 

```{r,eval=FALSE}
devtools::install_github("mdsumner/spdplyr")
```


# dplyr verbs for Spatial

Apply `dplyr` verbs to the attribute data of `sp` objects with dplyr verbs. 

See `?dplyr-Spatial'` for supported verbs. 


## How does it work? 

Most of it is easy, simply applying the analogous operation of the verb to the Spatial*DataFrame and subsetting/re-organizing as necessary. Some operations use the `spbabel::sptable` approach which provides a loss-less way to do what `ggplot2::fortify` does and its reverse. 


# Issues


**NOTE:** joins aren't really thought through or developed properly, send me your feedback / ideas / help!


