---
title: "spdplyr"
author: "Michael D. Sumner"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_width: 7
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{spdplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE,message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(maptools)
library(spdplyr)
```

# dplyr verbs for Spatial

Apply `dplyr` verbs to the attribute data of `sp` objects with dplyr verbs. 

See `?dplyr-Spatial'` for supported verbs. 

```{r}
data(quakes)
library(sp)
coordinates(quakes) <- ~long+lat
library(spdplyr)
## plot a subset of locations by number of stations
quakes %>% dplyr::filter(mag <5.5 & mag > 4.5) %>% dplyr::select(stations) %>% spplot()
```

We can use polygons and lines objects as well. 

```{r}
library(maptools)
data(wrld_simpl)
## put the centre-of-mass centroid on wrld_simpl as an attribute and filter/select
worldcorner <- wrld_simpl %>% 
  mutate(lon = coordinates(wrld_simpl)[,1], lat = coordinates(wrld_simpl)[,2]) %>% 
  filter(lat < -20, lon > 60) %>% 
  dplyr::select(NAME)

## demonstrate that we have a faithful subset of the original object
plot(worldcorner, asp = "")
text(coordinates(worldcorner), label = worldcorner$NAME, cex = 0.6)

worldcorner

## we can chain together standard operations as well as dplyr specific ones
wrld_simpl %>% as("SpatialLinesDataFrame") %>% summarise(big = max(AREA))
```

This approach is limited to the simple verbs  `arrange`, `distinct`, `filter`, `mutate`, `rename`, `select`, `slice`, `transmute`, and `summarize`. Summarize either on its own or after a `group_by` collapses to a non-topological union of all sub-geometries within a group without any consideration of topology or relationships. 

**BEWARE**: This is not a proven method for dealing with Spatial data, use at your own risk.

