---
title: "Raster extract"
author: "Michael Sumner"
date: "25 April 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Raster extract is slow for RasterLayer but not single-layer RasterBrick

```{r}
library(raster)
r <- disaggregate(raster(volcano), fact = 225)
if (fromDisk(r)) r <- readAll(r)
b <- brick(r)
if (fromDisk(b)) b <- readAll(b)

cells <- sample(ncell(r), 1e6)
# system.time(extract(r, cells))
#   user  system elapsed 
#   0.53    0.11    0.64 
# system.time(extract(b, cells))
#   user  system elapsed 
#   0.17    0.00    0.17 

```

## What about from disk? 

Extract assumes reading "line by line" - the rows of the raster as you seem them in a plot - when the raster is tiled this causes many un-necessary reading calls as each line is read from multiple tiles. 

```{r}
writeRaster(b, "notiling.tif", overwrite = TRUE )
gtiff0 <- brick("notiling.tif")
fromDisk(gtiff0)
class(gtiff0)

writeRaster(b, "tiling.tif", options = c("TILED=YES"), overwrite = TRUE )
gtiled <- brick("tiling.tif")
fromDisk(gtiled)
class(gtiled)
any(grepl("Block=256x256", system("C:/OsGeo4W64/bin/gdalinfo tiling.tif", intern = TRUE)))
library(rbenchmark)
benchmark(GeoTIFFLine = extract(gtiff0, cells), GeoTIFFTile = extract(gtiled, cells), replications = 1)
```



## What does this mean?

1. for GeoTIFF make sure you readAll from disk if it's tiled (use gdalinfo to find out, or gdalUtils package in R)
2.  use readAll(brick(raster(filename))) to get an in-memory brick


We can fix 1. with a change to the extract method if it can see that it's tiled, with 2. I don't know why this is yet. 

```{r}
benchmark(a0 = raster:::.cellValues(r, cells, 1), 
         a1 = raster:::.cellValues(b, cells, 1), 
         b1 = raster:::.cellValues(b[[1]], cells, 1), 
         replications = 1)

benchmark(a0 = raster:::.readCells(r, cells, 1), 
         a1 = raster:::.readCells(b, cells, 1), 
         b1 = raster:::.readCells(b[[1]], cells, 1), 
         replications = 1)



```